Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-04-21 ATVA 2025/experiments.tex
exp

\subsection{Comparison between different scoring functions.}

%The first group of experiments we run, tests the accuracy of choosing a set $Z$ of $K$ nodes using compensation strength, compared with randomly choosing the same number $K$ of nodes,  to understand if the choice is meaningful or any set $Z$ with the same size will result in a similar accuracy.
First, we measured the distance between SAS or GS and real improvement for all nodes in the first two layers of $5 \times 100$ wrt the accuracy of nodes in the third layer. On average, the distance using GS is 4 times larger than using SAS.

Then, to measure the impact of the scoring function to select neurons to open, 
we considered the largest complex CNN-B-Adv, with 20 000 nodes. 
We tested over the $\vx=85$th image in the CIFAR dataset.
To measure the accuracy, we measure the uncertainty of all nodes in the 4th layer:
the uncertainty of a node is the range between its computed lower and upper bound. 
We then average the uncertainty among all the nodes of the layer.
Formally, the uncertainty of a node $a$ with bounds $[\LB,\UB]$ is uncert$(a) = \UB(a) - \LB(a)$. The average uncertainty of layer $\ell$ is 
$\dfrac{\sum_{a\in l} \text{uncert}(a)}{size(\ell)}.$

@@

%We focus on the uncertainty of nodes in the third layers, wrt ReLU nodes in the first and second layer. The bounds for nodes of the first two layers are computed exactly using the full MILP encoding. 
We report in Figure \ref{fig_table3} the average uncertainty of $\MILP_X$ following the 
choice of the $K$ heaviest neurons of SAS, compared with a random choice, with Huang \cite{DivideAndSlide}, based on strength$(n) = (\UB(n)-\LB(n))\cdot |W_{nz}|$, and with GS.

@@

\begin{table}[t!]	
	\centering
	\begin{tabular}{|c||c c|c c|c c|c c|}
		\hline
		  Image 85& \multicolumn{2}{c|}{45 open ReLUs} & \multicolumn{2}{c|}{50 open ReLUs} 
		  & \multicolumn{2}{c|}{55 open ReLUs} & \multicolumn{2}{c|}{60 open ReLUs}\\ 
		 order & distance & time & distance & time & distance & time & distance & time \\
		\hline \hline
		Gurobi & 0.384 & 252s & 0.368&265  &0.342 & 287& 0.315  & 306s  \\ 
		SAS (static) & 0.384 & 448s & 0.368&759  &0.365 &971 & 0.355  & 971s \\
		GS (static) & 0.384 & {\bf 251s} &0.368 &{\bf 253}  & 0.342 &{\bf 270} & 0.315 & {\bf 281s} \\ \hline
	\end{tabular}
	\caption{Comparison of different orderings with selection of ReLUs by SAS.
	\vspace{-0.4cm}}
	\label{table.order}
\end{table}

SAS ordering is particularly counter-productive because it is accurate only for one branch (the most complicated one), whereas Gurobi can adapt the order to each branch.
However, the GS {\em ordering} (with the SAS {\em selection}) is better as it is general and not local to a solution, with better runtime than Gurobi, despite its staticness whereas Gurobi order is adaptative. Using SAS order for the solution branch, and GS ordering for other branches should be more efficient, but we cannot experiment that within the closed-source Gurobi solver.

@@

@@

\subsection{Finer Grain evaluation of Accuracy}

>>>>

\subsection{Comparison between different scoring functions (accuracy).}

%The first group of experiments we run, tests the accuracy of choosing a set $Z$ of $K$ nodes using compensation strength, compared with randomly choosing the same number $K$ of nodes,  to understand if the choice is meaningful or any set $Z$ with the same size will result in a similar accuracy.
%First, we measured the distance between SAS or GS and real improvement for all nodes in the first two layers of $5 \times 100$ wrt the accuracy of nodes in the third layer. On average, the distance using GS is 4 times larger than using SAS.

%Then, 
To measure the impact of the scoring function to select neurons to open, 
we considered the largest complex CNN-B-Adv, with 20 000 nodes. 
We tested over the $\vx=85$th image in the CIFAR dataset.
To measure the accuracy, we measure the uncertainty of all nodes in the 4th layer:
the uncertainty of a node is the range between its computed lower and upper bound. 
We then average the uncertainty among all the nodes of the layer.
Formally, the uncertainty of a node $a$ with bounds $[\LB,\UB]$ is uncert$(a) = \UB(a) - \LB(a)$. The average uncertainty of layer $\ell$ is 
$\dfrac{\sum_{a\in l} \text{uncert}(a)}{size(\ell)}.$

@@

%We focus on the uncertainty of nodes in the third layers, wrt ReLU nodes in the first and second layer. The bounds for nodes of the first two layers are computed exactly using the full MILP encoding. 
We report in Figure \ref{fig_table3} the average uncertainty of $\MILP_X$ following the 
choice of the $K$ heaviest neurons of SAS, compared with a random choice, with Huang \cite{DivideAndSlide}, based on strength$(n) = (\UB(n)-\LB(n))\cdot |W_{nz}|$, and with GS (here, $s_{FSB}$).

@@

\begin{table}[t!]	
	\centering
	\begin{tabular}{|c||c c|c c|c c|c c|}
		\hline
		  Image 85& \multicolumn{2}{c|}{45 open ReLUs} & \multicolumn{2}{c|}{50 open ReLUs} 
		  & \multicolumn{2}{c|}{55 open ReLUs} & \multicolumn{2}{c|}{60 open ReLUs}\\ 
		 order & distance & time & distance & time & distance & time & distance & time \\
		\hline \hline
		Gurobi & 0.384 & 252s & 0.368&265s  &0.342 & 287s& 0.315  & 306s  \\ 
		SAS (static) & 0.384 & 448s & 0.368&759s  &0.365 &971s & 0.355  & 971s \\
		GS (static) & 0.384 & {\bf 251s} &0.368 &{\bf 253s}  & 0.342 &{\bf 270s} & 0.315 & {\bf 281s} \\ \hline
	\end{tabular}
	\caption{Comparison of different orderings with selection of ReLUs by SAS.
	\vspace{-0.4cm}}
	\label{table.order}
\end{table}

SAS ordering is particularly counter-productive because it is accurate only for one branch (the most complicated one), whereas Gurobi can adapt the order to each branch.
However, the GS {\em ordering} (with the SAS {\em selection}) is better as it is general and not local to a solution, with better runtime than Gurobi, despite its staticness whereas Gurobi order is adaptative.
% Using SAS order for the solution branch, and GS ordering for other branches should be more efficient, but we cannot experiment that within the closed-source Gurobi solver.

@@

\subsection{Comparison with MILP (time and accuracy)}

Restricting the set $X$ of open ReLU nodes potentially hurts accuracy. 
Another strategy could be to still compute inductively accurate bounds for each nodes,
replacing pMILP ({\sf SAS, GS}, etc.) with a full MILP model, and stopping it early after some fixed time-out (100s,250s, $\ldots$, 2000s).

We evaluate in Figure \ref{fig555} the full MILP and different pMILP models on the output layer of CNN-B-Adv. The bounds for hidden layers have been computed with the same SAS method.
We compare the {\em distance to verify} the same Image 85, specifically the lower bound of an output neuron, the furthest away from verification.
%(the local robustness objective asks the difference between the value of the output neuron corresponding to the correct class and the value of any other output neuron to be positive). 
The curve is not as smooth as in Fig.~\ref{fig_table3}, as a unique neuron is considered rather than an average over neurons.

\begin{figure}[t!]
	\hspace*{-0.8cm}
	\includegraphics[scale=0.55]{CNN-B-ADV_layer7_comparison_image85_4methods.png}
	\caption{Distance to verify vs runtime: comparison between SAS, GS, full MILP and Huang's method for different number of ReLUs open / time-outs.}
	\label{fig555}
\end{figure}

{\sf SAS} (even {\sf GS}$=s_{FSB}$) is much more accurate than full MILP for every reasonable time-outs considered, with >10 times fastest runtime at better accuracy. The reason is that even the advanced MILP solver Gurobi struggles to sort out all ReLUs, and that is when having accurate scoring functions as {\sf SAS }is particularly important.

The difference of distance to verify between SAS and GS ISO number of ReLUs is very large. However, as the ReLUs opened by GS are not as important as the ReLUs opened by SAS, the runtime of GS ISO number of ReLUs is faster than for SAS. Still, the important metric here is runtime ISO accuracy (distance to verify), which is clearly favorable for SAS, close to 2x faster. Importantly, the accuracy gain by SAS is very predictable, with improvement with each open ReLUs, while it is much more erratic for GS. This helps setting a fixed number of ReLUs. Also, SAS reaches accuracy which would be hard to match by GS in number of ReLUs to open, needing an extremely high number of open ReLUs, and thus, a uncertain high runtime.

@@

\iffalse
\subsection{Finer Grain evaluation of Accuracy}
